---
title: "stlco_fimr_report"
author: "Michael Royce Tan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: false
header-includes:
  - \usepackage{helvet}
  - \usepackage[T1]{fontenc}
  - \renewcommand{\familydefault}{\sfdefault}
geometry: 
  - top = 0.8in
  - bottom = 0.5in
  - left = 0.5in
  - right = 0.5in
editor_options:
  chunk_output_type: console
---

<!-- ## r chuck global settings -->
```{r setup, include=FALSE}
# sets global default options for all code chunks
knitr::opts_chunk$set(
  echo = FALSE,            # hide code
  eval = TRUE,             # run code
  warning = FALSE,         # hide warnings
  message = FALSE,         # hide messages
  error = FALSE,           # stop rendering on errors
  fig.align = "center",    # center figures
  fig.width = 8,           # 
  fig.asp = 0.618,
  out.width = "\\textwidth",
  fig.path = "figs/",
  dpi = 300,               # high resolution figures
  tidy = FALSE,            # don't reformat code
  cache = FALSE,           # don't cache
  comment = "#>"           # comment prefix
  )
```


<!-- ### Pre-requisites -->
```{r, echo=FALSE, results='hide'}
# Load libraries
# library(tidyverse)
library(here)
library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(tidyr)
library(dplyr)
library(scales)
library(purrr)


```


<!-- # Importing and Saving Data -->
```{r, eval=FALSE, results='hide'}

# Importing data to R environment
ext_stlco_births <- read_csv(here::here("data", "stlco_birth.csv"))
ext_stlco_fetal_deaths <- read_csv(here::here("data", "stlco_fetal-death.csv"))

# create directory
dir.create(here::here("data", "raw"), recursive = TRUE)

# Saving data as rds
saveRDS(ext_stlco_births, here::here("data", "raw", "stlco_births.rds"))
saveRDS(ext_stlco_fetal_deaths, here::here("data", "raw", "stlco_fetal_deaths.rds"))

```


<!-- # Clean Up R environment after Saving -->
```{r, eval=FALSE, results='hide'}

#remove all objects from R environment
rm(list = ls())

# clean-up memory, trigger garbage collection
gc()
```


<!-- # Loading Raw Data -->
```{r, results='hide'}

#Loading rds data
stlco_births <- readRDS(here::here("data", "raw", "stlco_births.rds"))
stlco_fetal_deaths <- readRDS(here::here("data", "raw", "stlco_fetal_deaths.rds"))

```


<!-- # Explore Data -->
```{r, eval=FALSE, results='hide'}

# Explore stlco_births Data
glimpse(stlco_births)

# statistic summary
summary(stlco_births)

# Exploring unique values of each column
stlco_births |> 
  dplyr::distinct(year)
stlco_births |> 
  dplyr::distinct(sex)
stlco_births |> 
  dplyr::distinct(mom_race)

# Exploring mom_age 
invalid_mom_birth_age_count <- stlco_births |> 
  dplyr::filter(mom_age < 0) |>
  tally()
invalid_mom_birth_age_count

stlco_births |> 
  dplyr::filter(mom_age > 0 & mom_age < 12) |>
  tally()

# Explore stlco_births Data
glimpse(stlco_fetal_deaths)

summary(stlco_fetal_deaths)

stlco_fetal_deaths |> 
  dplyr::distinct(year)
stlco_fetal_deaths |> 
  dplyr::distinct(sex)
stlco_fetal_deaths |> 
  dplyr::distinct(mom_race)

# observations with mom birth age < 0
invalid_mom_birth_age_count

```


<!-- ## Checking Missing Data -->
```{r, eval=FALSE, results='hide'}


#Check for missing data in births
#stlco_births |> is.na() |> colSums()
#colSums(is.na(stlco_births))

missing_count_stlco_births <- stlco_births |> 
  dplyr::summarise(across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0)
missing_count_stlco_births

#Check for missing data in fetal deaths
#stlco_fetal_deaths |> is.na() |> colSums()
#colSums(is.na(stlco_births))

missing_count_stlco_fetal_deaths <- stlco_fetal_deaths |> 
  dplyr::summarise(across(everything(), ~ sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  dplyr::filter(n_missing > 0)
missing_count_stlco_fetal_deaths

```


\newpage
\vspace*{0.1in}
## Fetal Mortality Rate, Saint Louis County 2019 to 2023

\[
\text{Fetal mortality rate} =
\frac{\text{Fetal deaths at 20 weeks of gestation or more}}
{\text{Live births and fetal deaths at 20 weeks or more}}
\times 1{,}000
\]

\vspace*{0.3in}
For Normal Distribution (deaths $\ge$ 100):

\[
\begin{aligned}
\text{Lower: } & R_1 - 1.96 * R_1 * \frac{\text{RSE}(R_1)}{100} \\
\text{Upper: } & R_1 + 1.96 * R_1 * \frac{\text{RSE}(R_1)}{100}
\end{aligned}
\]

\[
\text{RSE of the fetal mortality rate:} \\[10pt]
\text{RSE} = 100 * \sqrt{\frac{1}{D} + \frac{1}{B}}
\]

\vspace*{0.3in}
For Poisson Distribution (deaths < 100):

\[
\begin{aligned}
\text{Lower: } & R_1 \cdot L(0.95, D_{\text{adj}}) \\
\text{Upper: } & R_1 \cdot U(0.95, D_{\text{adj}})
\end{aligned}
\]

\[
D_{\text{adj}} = \frac{D \bullet B}{D + B}
\]

Using the CDC Poisson L/U confidence-limit table for fetal and perinatal mortality rates.
Using qchisq in R, same values as the CDC Poisson L/U confidence-limit table

\[
\begin{aligned}
L &= \frac{\text{qchisq}(0.025, 2n)}{2n} \\[8pt]
U &= \frac{\text{qchisq}(0.975, 2(n+1))}{2n}
\end{aligned}
\]

```{r}
# Count total fetal deaths (20 weeks+)
total_fetal_deaths <- stlco_fetal_deaths |>
  dplyr::summarise(fetal_deaths = n())

# Count total live births
total_live_births <- stlco_births |>
  dplyr::summarise(live_births = n())


# Calculate fetal mortality rate
fmr_overall <- total_fetal_deaths |>
  dplyr::bind_cols(total_live_births) |>
  dplyr::mutate(
    total = fetal_deaths + live_births,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  )

#fmr_overall
```


<!-- ## Custom Functions for Calculating CI -->

```{r}
# Function for calculating Relative Standard Error, Normal Distribution
calculate_rse <- function(fetal_deaths, total) {
  rse <- 100 * sqrt((1/fetal_deaths) + (1/total))
}

# Function for adjusted fetal deaths for poisson distribution
calculate_adj_fetal_deaths <- function(fetal_deaths, total) {
  adj_fetal_deaths <- round((fetal_deaths * total) / (fetal_deaths + total))
}

# Function for Lower and Upper Limit multiplier for poisson distribution
# Using the CDC Poisson L/U confidence-limit table for fetal and perinatal mortality rates.
# Using qchisq, same values as the CDC Poisson L/U confidence-limit table
l_poisson <- function(n) {
  qchisq(0.025, 2*n) / (2*n)
}

u_poisson <- function(n) {
  qchisq(0.975, 2*(n+1)) / (2*n)
}
```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: overall_fetal_mortality_rate

fmr_overall_ci <- fmr_overall |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#fmr_overall_ci
```

```{r}
# Table using kable
# data: overall_fetal_mortality_rate_ci

table_fmr_overall_ci <- fmr_overall_ci |>
  select(fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total))

kable_fmr_overall_ci <- table_fmr_overall_ci |>
  knitr::kable(
    col.names = c(
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births and fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Fetal Mortality Rate, Saint Louis County, 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 3, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_fmr_overall_ci
#kable_fmr_overall_ci
```

\newpage
### Table 1: Overall Fetal Mortality Rate, Saint Louis County, 2019 to 2023
```{r table_01, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_fmr_overall_ci
```




<!-- ## Trend: Annual Fetal Mortality Rate, Saint Louis County 2019 to 2023 -->
```{r}

# Calculate annual fetal mortality rate, 2019 to 2023
fmr_annual <- stlco_fetal_deaths |>
  dplyr::count(year, name = "fetal_deaths") |>
  dplyr::left_join(
    stlco_births |> dplyr::count(year, name = "live_births"),
    by = "year"
  ) |>
  dplyr::mutate(
    total = live_births + fetal_deaths,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  )

#fmr_annual
```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: fmr_annual

fmr_annual_ci <- fmr_annual |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#fmr_annual_ci
```

```{r}
# Table using kable
# data: fmr_annual_ci

table_fmr_annual_ci <- fmr_annual_ci |>
  select(year,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total))

kable_fmr_annual_ci <- table_fmr_annual_ci |>
  knitr::kable(
    col.names = c(
      "Year",
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births and fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Annual Fetal Mortality Rate, Saint Louis County 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 4, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_fmr_annual_ci
#kable_fmr_annual_ci
```

### Table 2: Annual Fetal Mortality Rate, Saint Louis County, 2019 to 2023
```{r table_02, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_fmr_annual_ci
```

```{r}
# function for custom scale
size <- list(
    xxl = 12*1.618,
    xl = 11*1.618,
    lg = 10*1.618,
    md = 9*1.618,
    base = 7*1.618,
    sm = 6*1.618,
    xs = 5*1.618,
    xxs = 4*1.618,
    ss = 3*1.618)
```


```{r}
# Plot
# data: table_fmr_annual_ci

plot_fmr_annual_ci <- ggplot(table_fmr_annual_ci, aes(year, fmr_per_1000)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              alpha = 0.2, 
              fill = "#2c3e50") +
  geom_line(
    linewidth = 1.2,
    color = "#2c3e50") +
  geom_point(
    size = 3,
    color = "#2c3e50")+
  geom_text(
    aes(label = sprintf("%.2f", fmr_per_1000)),
    vjust = -1.5,
    hjust = -0.1) +
  scale_y_continuous(
    limits = c(0, 9),
    breaks = seq(0, 9, 2)) +
  labs(
    title = "Annual Fetal Mortality Rate, St. Louis County 2019-2023",
    subtitle = "Rate per 1,000 live births and fetal deaths",
    x = NULL,
    y = "FMR",
    caption = "Shaded area represents 95% confidence interval") +
  theme(
    plot.margin = margin(20, 20, 20, 20, "pt"),
    #axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.title = element_text(size = size$base),
    axis.text = element_text(size = size$sm),
    #legend.title = element_text(size = size$sm),
    #legend.text = element_text(size = size$sm),
    plot.title = element_text(size = size$lg,
                              margin = margin(b = 5)),
    plot.subtitle = element_text(size = size$base,
                                 margin = margin(b = 3)),
    plot.caption = element_text(hjust = 0,
                                size = size$sm),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.border = element_rect(
      color = "grey95",
      fill = NA,
      linewidth = 0.5),     # inside plot border
    panel.background = element_rect(
      color = NA,
      fill = "grey98"),    # only inside plot area
    plot.background = element_rect(
      color = "white",
      fill = NA,            # fill all plot area
      linewidth = 0)
  )
  
#plot_fmr_annual_ci

```

### Figure 1: Annual Fetal Mortality Rate, Saint Louis County, 2019 to 2023
```{r figure_01, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

plot_fmr_annual_ci
```



<!-- # Fetal Mortality Rate by Racial Groups -->
```{r}

# Calculate annual fetal mortality rate by race, Saint Louis County 2019 to 2023
fmr_race <- stlco_fetal_deaths |>
  dplyr::count(mom_race, name = "fetal_deaths") |>
  dplyr::left_join(
    stlco_births |> dplyr::count(mom_race, name = "live_births"),
    by = "mom_race"
  ) |>
  dplyr::mutate(
    total = live_births + fetal_deaths,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  ) |>
  dplyr::arrange(
    desc(fetal_deaths)
  )

#fmr_race

```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: fmr_race

fmr_race_ci <- fmr_race |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#fmr_race_ci

```


```{r}
# Tables using kable
# data: fmr_race_ci

table_fmr_race_ci <- fmr_race_ci |>
  select(mom_race,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total),
    mom_race = case_when(
      mom_race == "black" ~ "Black",
      mom_race == "white" ~ "White",
      mom_race == "other" ~ "Other",
      TRUE ~ "Missing/Unknown"),
    mom_race = factor(mom_race, 
                      levels = c("Black", "White", "Other", "Missing/Unknown")))

kable_fmr_race_ci <- table_fmr_race_ci |>
  knitr::kable(
    col.names = c(
      "Race",
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births and fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Fetal Mortality Rate by Race, Saint Louis County, 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 4, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_fmr_race_ci
#kable_fmr_race_ci

```

\newpage
### Table 3: Fetal Mortality Rate by Race, Saint Louis County, 2019 to 2023
```{r table_03, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_fmr_race_ci
```

<!-- ## Trend:Annual FMR by Racial Groups -->

```{r}
# Calculate annual fetal mortality rate by race, Saint Louis County 2019 to 2023
trend_fmr_race <- stlco_fetal_deaths |>
  dplyr::count(year, mom_race, name = "fetal_deaths") |>
  dplyr::left_join(
    stlco_births |> dplyr::count(year, mom_race, name = "live_births"),
    by = c("year", "mom_race")
  ) |>
  dplyr::mutate(
    total = live_births + fetal_deaths,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  ) |>
  dplyr::arrange(
    year,
    desc(fetal_deaths)
  )

#trend_fmr_race
```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: trend_fmr_race

trend_fmr_race_ci <- trend_fmr_race |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#trend_fmr_race_ci
```

```{r}
# Tables using Kable
# data: trend_fmr_race_ci

table_trend_fmr_race_ci <- trend_fmr_race_ci |>
  select(year,
         mom_race,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total),
    mom_race = case_when(
      mom_race == "black" ~ "Black",
      mom_race == "white" ~ "White",
      mom_race == "other" ~ "Other",
      TRUE ~ "Missing/Unknown"),
    mom_race = factor(mom_race, 
                      levels = c("Black", "White", "Other", "Missing/Unknown")))


kable_trend_fmr_race_ci <- table_trend_fmr_race_ci |>
  knitr::kable(
    col.names = c(
      "Year",
      "Race",
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births/fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Fetal Mortality Rate by Race, Saint Louis County, 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 5, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_trend_fmr_race_ci
#kable_trend_fmr_race_ci
```

### Table 4: Fetal Mortality Rate by Race, Saint Louis County, 2019 to 2023
```{r table_04, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_trend_fmr_race_ci
```

```{r}
#data: table_trend_fmr_race_ci

# Filter out Missing/Unknown for ggplot
table_trend_fmr_race_ci_drop <- table_trend_fmr_race_ci |>
  dplyr::filter(mom_race != "Missing/Unknown")
```

```{r}
# Plot 1
# data: table_trend_fmr_race_ci_drop

plot_table_trend_fmr_race_ci_drop <- table_trend_fmr_race_ci_drop |>
  ggplot(
    aes(x = year, y = fmr_per_1000,
        group = mom_race,
        color = mom_race, fill = mom_race)) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper), 
    alpha = 0.15, 
    color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(
    limits = c(0, 15),
    breaks = seq(0, 15, 3)) +
  scale_color_manual(values = c(
    "Black" = "#8FBF87",
    "White" = "#4A6FA5",
    "Other" = "#F28C6B"),
    name = "Maternal Race") +
  scale_fill_manual(values = c(
    "Black" = "#8FBF87",
    "White" = "#4A6FA5",
    "Other" = "#F28C6B"),
    name = "Maternal Race") +
  labs(
    title = "Annual Fetal Mortality Rate by Maternal Race, St. Louis County, 2019-2023",
    subtitle = "Rate per 1,000 live births and fetal deaths",
    x = NULL,
    y = "FMR",
    caption = "Shaded area represents 95% confidence interval") +
  theme(
    plot.margin = margin(20, 20, 20, 20, "pt"),
    #axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.title = element_text(size = size$base),
    axis.text = element_text(size = size$sm),
    legend.title = element_text(size = size$sm),
    legend.text = element_text(size = size$sm),
    legend.position = "top",
    plot.title = element_text(size = size$lg,
                              margin = margin(b = 5)),
    plot.subtitle = element_text(size = size$base,
                                 margin = margin(b = 3)),
    plot.caption = element_text(hjust = 0,
                                size = size$sm),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.border = element_rect(
      color = "grey95",
      fill = NA,
      linewidth = 0.5),     # inside plot border
    panel.background = element_rect(
      color = NA,
      fill = "grey98"),    # only inside plot area
    plot.background = element_rect(
      color = "white",
      fill = NA,            # fill all plot area
      linewidth = 0)
  )

#plot_table_trend_fmr_race_ci_drop
```

\newpage
### Figure 2: Annual Fetal Mortality Rate by Race, Saint Louis County, 2019 to 2023
```{r figure_02, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

plot_table_trend_fmr_race_ci_drop
```


<!-- # Fetal Mortality Rate by Sex -->
```{r}
# Calculate annual fetal mortality rate by Sex, Saint Louis County 2019 to 2023
fmr_sex <- stlco_fetal_deaths |>
  dplyr::count(sex, name = "fetal_deaths") |>
  dplyr::left_join(
    stlco_births |> dplyr::count(sex, name = "live_births"),
    by = "sex"
  ) |>
  dplyr::mutate(
    total = live_births + fetal_deaths,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  ) |>
  dplyr::arrange(
    desc(fetal_deaths)
  )

#fmr_sex

```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: fmr_sex

fmr_sex_ci <- fmr_sex |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#fmr_sex_ci

```

```{r}
# Tables using Kable
# data: fmr_sex_ci

table_fmr_sex_ci <- fmr_sex_ci |>
  select(sex,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total)) 


kable_fmr_sex_ci <- table_fmr_sex_ci |>
  knitr::kable(
    col.names = c(
      "Sex",
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births and fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Fetal Mortality Rate by Sex, Saint Louis County, 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 4, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_fmr_sex_ci
#kable_fmr_sex_ci
```

\newpage
### Table 5: Fetal Mortality Rate by Sex, Saint Louis County, 2019 to 2023
```{r table_05, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_fmr_sex_ci
```


<!-- ## Trend: Annual FMR by Sex -->
```{r}
# Calculate annual fetal mortality rate by Sex, Saint Louis County 2019 to 2023
trend_fmr_sex <- stlco_fetal_deaths |>
  dplyr::count(year, sex, name = "fetal_deaths") |>
  dplyr::left_join(
    stlco_births |> dplyr::count(year, sex, name = "live_births"),
    by = c("year", "sex")
  ) |>
  dplyr::mutate(
    total = live_births + fetal_deaths,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  ) |>
  dplyr::arrange(
    year,
    desc(fetal_deaths)
  )

#trend_fmr_sex
```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: trend_fmr_sex

trend_fmr_sex_ci <- trend_fmr_sex |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#trend_fmr_sex_ci
```

```{r}
# Tables using Kable
# data: trend_fmr_sex_ci

table_trend_fmr_sex_ci <- trend_fmr_sex_ci |>
  select(year,
         sex,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total))

kable_trend_fmr_sex_ci <- table_trend_fmr_sex_ci |>
  knitr::kable(
    col.names = c(
      "Year",
      "Sex",
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births and fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Annual Fetal Mortality Rate by Sex, Saint Louis County 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 5, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_trend_fmr_sex_ci
#kable_trend_fmr_sex_ci
```

### Table 6: Annual Fetal Mortality Rate by Sex, Saint Louis County, 2019 to 2023
```{r table_06, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_trend_fmr_sex_ci
```

```{r}
# as numeric for ggplot
#data: trend_fmr_sex_ci

table_trend_fmr_sex_ci_num <- trend_fmr_sex_ci |>
  select(year,
         sex,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::filter(sex != "U") |>
  dplyr::mutate(
    sex_label = case_when(
      sex == "M" ~ "Male",
      sex == "F" ~ "Female"),
    sex_label = factor(sex_label, 
                      levels = c("Male", "Female")))
```



```{r}
#plot
# data: table_trend_fmr_sex_ci_num

plot_table_trend_fmr_sex_ci_num <- table_trend_fmr_sex_ci_num |>
  ggplot(aes(x = year, y = fmr_per_1000, color = sex_label, fill = sex_label, group = sex_label)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              alpha = 0.15,
              color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
    "Male" = "#8FBF87",
    "Female" = "#F28C6B"),
    name = "Sex") +
  scale_fill_manual(values = c(
    "Male" = "#8FBF87",
    "Female" = "#F28C6B"),
    name = "Sex") +
  labs(
    title = "Annual Fetal Mortality Rate by Sex, St. Louis County, 2019-2023",
    subtitle = "Rate per 1,000 live births and fetal deaths",
    x = NULL,
    y = "FMR",
    caption = "Shaded area represents 95% confidence interval") +
  theme(
    plot.margin = margin(20, 20, 20, 20, "pt"),
    #axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.title = element_text(size = size$base),
    axis.text = element_text(size = size$sm),
    legend.title = element_text(size = size$sm),
    legend.text = element_text(size = size$sm),
    legend.position = "top",
    plot.title = element_text(size = size$lg,
                              margin = margin(b = 5)),
    plot.subtitle = element_text(size = size$base,
                                 margin = margin(b = 3)),
    plot.caption = element_text(hjust = 0,
                                size = size$sm),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.border = element_rect(
      color = "grey95",
      fill = NA,
      linewidth = 0.5),     # inside plot border
    panel.background = element_rect(
      color = NA,
      fill = "grey98"),    # only inside plot area
    plot.background = element_rect(
      color = "white",
      fill = NA,            # fill all plot area
      linewidth = 0)
  )

#plot_table_trend_fmr_sex_ci_num
```


### Figure 3: Annual Fetal Mortality Rate by Sex, Saint Louis County, 2019 to 2023

```{r figure_03, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

plot_table_trend_fmr_sex_ci_num
```



<!-- # Fetal Mortality Rate by Maternal Age Groups -->
```{r}
# Function for creating maternal age groups
create_age_groups <- function(age) {
  dplyr::case_when(
    age < 15 ~ "Under 15",
    age >= 15 & age <= 19 ~ "15-19",
    age >= 20 & age <= 24 ~ "20-24",
    age >= 25 & age <= 29 ~ "25-29",
    age >= 30 & age <= 34 ~ "30-34",
    age >= 35 & age <= 39 ~ "35-39",
    age >= 40 & age <= 44 ~ "40-44",
    age >= 45 ~ "45+",
    TRUE ~ NA_character_
  )
}
```

```{r}
# Calculate fetal mortality rate by maternal age group, Saint Louis County 2019 to 2023
fmr_age_group <- stlco_fetal_deaths |>
  dplyr::mutate(age_group = create_age_groups(mom_age)) |>
  dplyr::count(age_group, name = "fetal_deaths") |>
  dplyr::left_join(
    stlco_births |> 
      dplyr::mutate(age_group = create_age_groups(mom_age)) |>
      dplyr::count(age_group, name = "live_births"),
    by = "age_group"
  ) |>
  dplyr::mutate(
    total = live_births + fetal_deaths,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  ) 

#fmr_age_group
```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: fmr_age_group

fmr_age_group_ci <- fmr_age_group |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#fmr_age_group_ci
```

```{r}
# Tables using Kable
# data: fmr_age_group_ci

table_fmr_age_group_ci <- fmr_age_group_ci |>
  select(age_group,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total))

kable_fmr_age_group_ci <- table_fmr_age_group_ci |>
  knitr::kable(
    col.names = c(
      "Age Group",
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births and fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Fetal Mortality Rate by Maternal Age Group, Saint Louis County 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 4, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_fmr_age_group_ci
#kable_fmr_age_group_ci
```

\newpage
### Table 7: Fetal Mortality Rate by Maternal Age Group, Saint Louis County, 2019 to 2023
```{r table_07, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_fmr_age_group_ci
```


<!-- ## Trend: Annual FMR by Maternal Age -->
```{r}
# Calculate annual fmr by maternal age group, Saint Louis County 2019 to 2023
trend_fmr_age_group <- stlco_fetal_deaths |>
  dplyr::mutate(age_group = create_age_groups(mom_age)) |>
  dplyr::count(year, age_group, name = "fetal_deaths") |>
  dplyr::left_join(
    stlco_births |> 
      dplyr::mutate(age_group = create_age_groups(mom_age)) |>
      dplyr::count(year, age_group, name = "live_births"),
    by = c("year", "age_group")
  ) |>
  dplyr::mutate(
    total = live_births + fetal_deaths,
    fmr_per_1000 = (fetal_deaths / total) * 1000
  ) 

#trend_fmr_age_group
```

```{r}
# Calculating 95% CI 
# using Poisson Distribution (deaths < 100) or Normal Distribution (deaths >= 100)
# data: trend_fmr_age_group

trend_fmr_age_group_ci <- trend_fmr_age_group |>
  mutate(
    adj_fetal_deaths = calculate_adj_fetal_deaths(fetal_deaths, total),
    lower_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * l_poisson(adj_fetal_deaths)),
    upper_poisson = case_when(
      adj_fetal_deaths < 1 ~ NA_real_,
      TRUE ~ fmr_per_1000 * u_poisson(adj_fetal_deaths)),
    rse = calculate_rse(fetal_deaths, total),
    lower_normal = fmr_per_1000 - (1.96 * fmr_per_1000 * rse/100),
    upper_normal = fmr_per_1000 + (1.96 * fmr_per_1000 * rse/100),
    lower = dplyr::case_when(
      fetal_deaths >= 100 ~ lower_normal,
      TRUE ~ lower_poisson),
    upper = dplyr::case_when(
      fetal_deaths >= 100 ~ upper_normal,
      TRUE ~ upper_poisson),
    ci_method = dplyr::case_when(
      fetal_deaths >= 100 ~ "Normal",
      TRUE ~ "Poisson")
  )

#trend_fmr_age_group_ci
```

```{r}
# Tables using Kable
# data: fmr_age_group_ci

table_trend_fmr_age_group_ci <- trend_fmr_age_group_ci |>
  select(year,
         age_group,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    live_births = scales::comma(live_births),
    total = scales::comma(total))

kable_trend_fmr_age_group_ci <- table_trend_fmr_age_group_ci |>
  knitr::kable(
    col.names = c(
      "Year",
      "Age Group",
      "Fetal Deaths",
      "Live Births",
      "Total",
      "per 1000 live births and fetal deaths",
      "Lower",
      "Upper",
      "CI Method"),
    digits = 2,
    caption = "Fetal Mortality Rate by Maternal Age Group, Saint Louis County 2019 to 2023") |>
  kableExtra::add_header_above(c(" " = 5, "Fetal Mortality Rate" = 1, "95% CI" = 2, " " = 1)) |>
  kableExtra::kable_styling(latex_options = c("HOLD_position", "scale down"), 
                full_width = FALSE,
                position = "center")

#table_trend_fmr_age_group_ci
#kable_trend_fmr_age_group_ci
```


## Table 8: Annual Fetal Mortality Rate by Maternal Age Group, Saint Louis County, 2019 to 2023
```{r table_08, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

kable_trend_fmr_age_group_ci
```

```{r}
# table with live births and total as numeric for ggplot
# data: trend_fmr_age_group_ci

table_trend_fmr_age_group_ci_num <- trend_fmr_age_group_ci |>
  select(year,
         age_group,
         fetal_deaths,
         live_births,
         total,
         fmr_per_1000,
         lower,
         upper,
         ci_method) |>
  dplyr::mutate(
    age_group = factor(age_group, 
                       levels = c("15-19", "20-24", "25-29", "30-34", "35-39", "40-44")))
```


```{r}
#Plot 
# data: table_trend_fmr_age_group_ci_num or table_trend_fmr_age_group_ci

plot_table_trend_fmr_age_group_ci <- table_trend_fmr_age_group_ci |>
  ggplot(aes(x = year, y = fmr_per_1000)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              alpha = 0.2,
              fill = "#2c3e50") +
  geom_line(
    linewidth = 0.8,
    color = "#2c3e50") +
  geom_point(
    size = 1.5,
    color = "#2c3e50") +
  #geom_text(
    #aes(label = sprintf("%.2f", fmr_per_1000)),
    #vjust = -0.8) +
  facet_wrap(
    ~ age_group,
    ncol = 3, 
    scales = "fixed") +
  scale_x_continuous(
    breaks = c(2019, 2021, 2023)) +
  labs(
    title = "Fetal Mortality Rate by Maternal Age Group, St. Louis County, 2019-2023",
    subtitle = "Rate per 1,000 live births and fetal deaths",
    x = NULL,
    y = "FMR",
    caption = "Shaded area represents 95% confidence interval") +
  theme(
    plot.margin = margin(20, 20, 20, 20, "pt"),
    #axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.title = element_text(size = size$base),
    axis.text = element_text(size = size$sm),
    legend.title = element_text(size = size$sm),
    legend.text = element_text(size = size$sm),
    legend.position = "top",
    plot.title = element_text(size = size$lg,
                              margin = margin(b = 5)),
    plot.subtitle = element_text(size = size$base,
                                 margin = margin(b = 3)),
    plot.caption = element_text(hjust = 0,
                                size = size$sm),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.border = element_rect(
      color = "grey95",
      fill = NA,
      linewidth = 0.5),     # inside plot border
    panel.background = element_rect(
      color = NA,
      fill = "grey98"),    # only inside plot area
    plot.background = element_rect(
      color = "white",
      fill = NA,            # fill all plot area
      linewidth = 0)
  )
  
#plot_table_trend_fmr_age_group_ci
```

\newpage
## Figure 4: Annual Fetal Mortality Rate by Maternal Age Group, Saint Louis County, 2019 to 2023
```{r figure_04, fig.width=8, fig.asp=0.618, fig.align='center', out.width="\\textwidth"}

plot_table_trend_fmr_age_group_ci
```


# Reference

1. Centers for Disease Control and Prevention. (2023). User guide to the 2023 fetal death public use file. https://ftp.cdc.gov/pub/Health_Statistics/NCHS/Dataset_Documentation/DVS/fetaldeath/2023fetaluserguide.pdf

2. MacDorman, M. F., & Gregory, E. C. W. (2015, July 23). Fetal and perinatal mortality: United States, 2013 (National Center for Health Statistics. Vital and Health Statistics, 64[8], 1–24). https://stacks.cdc.gov/view/cdc/231851


